New lower bounds for Schur and weak Schur numbers

### 2021-04-06 article/article.tex
Add files via upload

@@

>>>>

\qquad In this section, we use Rowley's constructions [7] in a Schur context. To improve lower bounds for Ramsay's numbers, Rowley introduces partitions verifying some properties which can be extended using method which generalizes Abbott-Hanson's [8] construction. Rolwey named these partitions "templates", and we will keep this name in the entire article. Finding suitable templates and applying his method to these will allow us to find new lower bounds for Schur methods.

\subsection{Definition of SE\(_+\)}

\begin{definition}
Let \( n \in \mathbb{N}^*\), there exists a greatest integer that we note \(SE_+(n)\) (resp. \(SE_-(n)\)) such that \( [\![1, SE_+(n)]\!]\) (resp. \( [\![1, SE_-(n)]\!]\)) can be partitionned (resp. can not be partionned) into \(n\) sum-free subsets \(A_1, A_2, ..., A_n\) which verify :
\[
\forall k \in [\![1, n-1]\!], \forall (x,y) \in A_k^2, x+y > SE_+(n)
\Longrightarrow x+y-SE_+(n) \notin A_k
\]
\end{definition}
We notice that in general, unlike regular Schur numbers, we don't have \(SE_+(n) + 1 \neq SE_-(n)\)

\begin{theorem}
Let \(n \in [\![2, +\infty]\!]\), we have :
\[
2S(n-1)+1 \leqslant SE_+(n) \leqslant S(n)
\]
\end{theorem}

\subsection{}

\subsection{}

@@

In this section, we generalize Rowley's constructions in [2]. We then introduce, by analogy with the third section, the integer \(WE_+(n)\) to build suitable templates.

\subsection{Lower bound for Weak Schur numbers using Schur and Weak Schur numbers}

The following theorem, inspired by Rowley's work, was found and proved by Romain Ageron.
\begin{theorem}
Let $(n,p), (r,s) \in (\mathbb{N}^*)^2$. If there exists a partition of $r$ weakly sum-free subsets of $[\![1,n]\!]$ and a partition of $s$ sum-free subsets of $[\![1,p]\!]$ then there exists a partition of $r+s$ weakly sum-free subsets of $[\![1,p(n+\left \lceil \frac{n}{2} \right \rceil + 1)+n]\!]$
\end{theorem}
In particular, if we choose $n = WS(r)$ and $p = WS(s)$ in the last theorem, the next corollary follows.
\begin{corollary}
$ \forall (n,m) \in (\mathbb{N}^*)^2 \text{, } WS(n+m) \geqslant S(m) \left (WS(n) + \left \lceil \frac{WS(n)}{2} \right \rceil +1 \right) + WS(n)$
\end{corollary}
\textsc{Proof :} Let $(n,p), (r,s) \in (\mathbb{N}^*)^2$,  $N = p(n+\left \lceil \frac{n}{2} \right \rceil + 1)+n$, $\alpha = \left \lceil \frac{n}{2} \right \rceil > 0$ and $\beta = n + \alpha + 1$. We denote by $f$ the projection of the equivalence relation induced by the partition of $[\![1,n]\!]$ and $g$ the one induced by the partition of $[\![1,p]\!]$. Each equivalence class is represented by a single integer, therefore :
\[ f : [\![1,n]\!] \longrightarrow [\![1,r]\!] \text{ and } \forall (x,y) \in [\![1,n]\!]^2, \left\{
    \begin{array}{ll}
        x \neq y \\
        f(x) = f(y)
    \end{array}
\right.
\Longrightarrow f(x+y) \neq f(x)
\]
\[g : [\![1,p]\!] \longrightarrow [\![1,s]\!] \text{ and } \forall (x,y) \in [\![1,n]\!]^2 \text{, } f(x) = f(y) \Longrightarrow f(x+y) \neq f(x)
\]
Let us start by parting the integers of $[\![1,N]\!]$ in two subsets $\mathcal{A}$ and $\mathcal{B}$ where $\mathcal{A} = [\![1,\alpha]\!] \cup \{a\beta + u \mid (a,u) \in [\![0,p]\!] \times [\![\alpha + 1,n]\!]\}$ and $\mathcal{B} = \{a\beta + u \mid (a,u) \in [\![1,p]\!] \times [\![-\alpha,\alpha]\!]\}$.\\
\\
First, \underline{$\mathcal{A} \cap \mathcal{B} = \varnothing$} : \\
Suppose there exists $x \in \mathcal{A} \cap \mathcal{B} = \varnothing$. Then there are $(a,u) \in [\![0,p]\!] \times [\![\alpha + 1,n]\!]$ and $(b,v) \in [\![1,p]\!] \times [\![-\alpha,\alpha]\!]$ such that $x = a\beta + u = b\beta +v$. By definition of $\alpha$ and  $\beta$ we have $u \in [\![\alpha + 1,n]\!] \subset [\![0,\beta - 1]\!]$. From there, we distinguish two cases :
\begin{itemize}
\item If $v \in [\![0,\alpha]\!]$ then $v \in [\![0,\beta - 1]\!]$ and $v \neq u$ because $v < \alpha + 1 \leqslant u$
\item If $v \in [\![-\alpha,-1]\!]$, we note $\tilde{v} = \beta + v$ and thus have $x = (b-1)\beta + \tilde{v}$ with $\tilde{v} \in [\![\beta - \alpha,\beta - 1]\!] \subset [\![0,\beta - 1]\!]$ and $\tilde{v} \neq u$ because $u < n+1 = \beta - \alpha \leqslant \tilde{v}$.
\end{itemize}
In either cases, we run into a contradiction because of the remainder's uniqueness in the euclidean division of $x$ by $\beta$.\\
\\
Then, we have \underline{$\mathcal{A} \cup \mathcal{B} = [\![1,N]\!]$}:
\begin{itemize}
\item On one hand : $1 = \text{min($\mathcal{A}$)} \leqslant \text{max($\mathcal{A}$)} = p\beta + n = N$ and $1 \leqslant \beta - \alpha = \text{min($\mathcal{B}$)} \leqslant \text{max($\mathcal{B}$)} = p\beta + \alpha \leqslant N$, which gives $\mathcal{A} \cup \mathcal{B} \subset [\![1,N]\!]$.
\item On the other hand, let $x \in  [\![1,N]\!]$. If $x \leqslant \alpha$, we directly have $x \in \mathcal{A}$, let us then suppose that $x > \alpha$ and write $x = a\beta + u$ the euclidean division of $x$ by $\beta$. We have $x \leqslant N$, thus $a \leqslant p$. We distinguish three cases : \\
- If $u \in [\![0,\alpha]\!]$ then we necessarily have $a \geqslant 1$ because $x > \alpha$, and so $x \in \mathcal{B}$.\\
- If $u \in [\![\alpha + 1,n]\!]$, then $x \in \mathcal{A}$. \\
- If $u \in [\![n + 1,\beta - 1]\!]$ then $x = (a+1)\beta - (\beta - u)$ with $-\alpha \leqslant \beta - u \leqslant 0$. Furthermore, $a \leqslant p - 1$, else we would have $x > N$, and so $x \in \mathcal{B}$ \\
In any case, $x \in \mathcal{A} \cup \mathcal{B}$ and we can thus conclude that $[\![1,N]\!] \subset \mathcal{A} \cup \mathcal{B}$.
\end{itemize}
This first partition of $[\![1,N]\!]$ will help us to define our final partition by the projection of its equivalence relation. We thereby define $h : [\![1,N]\!] \longrightarrow [\![1,r+s]\!]$ as such :\\
- If $x \in \mathcal{A}$ then $h(x) = f(x \text{ mod } \beta)$ (well defined because $x \text{ mod } \beta \in [\![1,N]\!]$)\\
- If $x \in \mathcal{B}$ then $x = a\beta + u$ with a unique $(a,u) \in [\![1,p]\!] \times [\![-\alpha,\alpha]\!]$ and we define $h(x) = r + g(a)$\\
The fact that $(\mathcal{A}, \mathcal{B})$ is a partition of $[\![1,N]\!]$ ensures that this definition of $h$ is valid. We then have to verify that $h$ induces weakly sum-free subsets.\\
\\
\underline{The classes of equivalence $h(x)$ for $x \in \mathcal{A}$ are weakly sum-free :}
\\
\\
Let $(x,y) \in \mathcal{A}^2$ such that $h(x) = h(y)$, $x \neq y$ and $x + y \leqslant N$
\begin{itemize}
\item If $(x,y) \in [\![1,\alpha]\!]^2$ :\\
We have $x + y \leqslant 2\alpha \leqslant n$ and $x + y = 0\beta + x + y$, therefore $x + y \in \mathcal{A}$. Then, by definition : $h(x) = f(x)$, $h(y) = f(y)$ and $h(x+y) = f(x+y)$, which gives us, thanks to the property verified by $f$, that $h(x+y) \neq h(x)$.
\item If $(x,y) \in [\![1,\alpha]\!] \times ( \mathcal{A} \text{ \textbackslash} \text{ } [\![1,\alpha]\!] )$ :\\
We write $y = a\beta + u$ with $(a,u) \in [\![0,p]\!] \times [\![\alpha + 1,n]\!]$. Then $x+y = a\beta + x + u = (a+1)\beta + x + u - \beta$, and if $x + u > n$ it follows that $a \leqslant p-1$ since $x+y \leqslant N$, and $-\alpha \leqslant x + u - \beta \leqslant -1$. Therefore $x+y \in \mathcal{B}$ and $h(x+y) \neq h(x) = f(x)$ by definition of h. On the contrary, if $x - u \leqslant n$, then $x+y \in \mathcal{A}$ and $h(x+y) = f(x+u)$ because $x+u$ is actually the remainder of the euclidean division of $x+y$ by $\beta$. Moreover, $h(x) = f(x)$, $x < u$ and, with our initial hypothesis, $h(x) = h(y) = f(u)$. The property verified by $f$ gives us $f(x+u) \neq f(x)$ which can be rewritten as $h(x+y) \neq h(x)$.
\item If $(x,y) \in ( \mathcal{A} \text{ \textbackslash} \text{ } [\![1,\alpha]\!] ) \times [\![1,\alpha]\!]$ : \\
This case is handled exactly like the previous one by swaping the roles of $x$ and $y$.
\item If $(x,y) \in ( \mathcal{A} \text{ \textbackslash} \text{ } [\![1,\alpha]\!] )^2$ : \\
We write $x = a\beta + u$ and $y = b\beta + v$ with $(a,u)$ and $(b,v)$ in $[\![0,p]\!] \times [\![\alpha + 1,n]\!]$. Then $x+y = (a+b)\beta + u+v = (a+b+1)\beta + u + v - \beta$ with $a+b \leqslant p-1$ (else we would have $x+y > N$ because $u+v > n$) and $-\alpha \leqslant u + v - \beta \leqslant \alpha$, therefore $x+y \in \mathcal{B}$ and by definition $h(x+y) \neq h(x)$.
\end{itemize}
In any case, $h(x+y) \neq h(x)$ and the classes of equivalence $h(x)$ for $x \in \mathcal{A}$ are weakly sum-free.\\
\\
\underline{The classes of equivalence $h(x)$ for $x \in \mathcal{B}$ are weakly sum-free :}
\\
\\
Let $(x,y) \in \mathcal{B}^2$ such that $h(x) = h(y)$, $x \neq y$ and $x + y \leqslant N$.\\
We write $x = a\beta + u$ and $y = b\beta + v$ with $(a,u)$ and $(b,v)$ in $[\![1,p]\!] \times [\![-\alpha,\alpha]\!]$. We have $h(x) = r + g(a)$ and $h(y) = r + g(b)$, therefore $g(a) = g(b)$. We also have $x+y = (a+b)\beta + u +v$.\\
If $u + v \in [\![-\alpha,\alpha]\!]$, then $x+y \in \mathcal{B}$ and $h(x+y) = g(a+b)$, hence we can deduce that $h(x+y) \neq h(x)$ because of the property verified by $g$. On the contrary, if $u+v \notin [\![-\alpha,\alpha]\!]$, then necessarily $x+y \in \mathcal{A}$. Suppose $x+y \in \mathcal{B}$, then $x+y = c\beta + w$ with $(c,w) \in [\![1,p]\!] \times [\![-\alpha,\alpha]\!]$. Thus, $c\beta + w = (a+b)\beta + u + v$ and $(a+b-c)\beta = w-u-v$. Furthermore $a+b-c \neq 0$, else we would have $u+v = w \in [\![-\alpha,\alpha]\!]$. This finally leads to the following inequality :
\[\beta \leqslant |a+b-c|\beta = |w-u-v| \leqslant |w| + |u| + |v| \leqslant 3\alpha \leqslant n + \alpha < \beta
\]
which is absurd. We can therefore conclude that $x+y \in \mathcal{A}$ and by definition of $h$, $h(x+y) \neq h(x)$, proving that the classes of equivalence $h(x)$ for $x \in \mathcal{B}$ are weakly sum-free.\\
\\
Finally, we have showed that every classe of equivalence induced by $h$ is weakly sum-free, which ends the proof.
